---
title: "Homework7"
author: "Kitty Harris"
output: 
  pdf_document:
    includes:
      in_header: "input/preamble.tex"
---

```{r include=FALSE}
#imports
library(rstan); library(bayesutils); library(coda); 
library(ggplot2); library(bayesplot); 

#for use later
n = length(bank_salary$bsalary) #numentries
v = var(bank_salary$bsalary)    #sample variance

#stan
stanmod = "
data {
  int<lower=1>  n;    //num obs
  vector[n]     bsal; //main data
  vector[n]     male; //covariate
  vector[n]     educ; //covariate
  vector[n]     xp;   //covariate
  vector[n]     time; //covariate
  real<lower=0> v;    //sample variance
}
parameters{
  real<lower=0> sigs; //variance
  real          bet0; //regression coefficients vvv
  real          bet1;
  real          bet2;
  real          bet3;
  real          bet4;
}
model{
  //static priors* used by all models
  sigs ~ inv_gamma(0.01,0.01);//variance
  for(i in 1:n){              //data distribution
    bsal[i] ~ normal(bet0 + bet1*male[i] + bet2*educ[i] + bet3*xp[i] + bet4*time[i],sigs^(1./2));
  }
  
  //priors for regression coefficients; these change each time
  bet0 ~ normal(0,sqrt(1002)); 
  bet1 ~ normal(0,sqrt(1002));
  bet2 ~ normal(0,sqrt(1002));
  bet3 ~ normal(0,sqrt(1002));
  bet4 ~ normal(0,sqrt(1002));
}
generated quantities {
  real Rbsq;
  Rbsq = 1 - sigs / v;
}"

#Format data for stan
stan_dat = list(n = n, v = v, bsal = bank_salary$bsalary,
                male = bank_salary$male, educ = bank_salary$education, 
                xp = bank_salary$experience, time = bank_salary$time)

if (!file.exists("stanmodels/hw7q1.rda")){ #Generate the model if it is missing.
  mod = stan_model(model_code = stanmod)
  save(mod,file="stanmodels/hw7q1.rda",compress="xz")
}
load(file = "stanmodels/hw7q1.rda")
fit = sampling(mod,data=stan_dat,iter=100000,chains=4,warmup = 50000,
               show_messages = FALSE);

X = cbind(1, bank_salary$male, bank_salary$education, bank_salary$experience, 
          bank_salary$time)
stanmod_zellner = "
data {
  int<lower=1>  n;    //num obs
  vector[n]     bsal; //main data
  real<lower=0> csq;  //zellner's g-prior var
  matrix[n,5]   X;    //covariates
  vector[5]     mu0;  //starting means for covariants
  cov_matrix[5] V;    //covariance matrix
  real<lower=0> v;    //sample variance
}
parameters{
  real<lower=0> sigs; //variance
  vector[5]     beta; //regression coefficients
}
transformed parameters{
  vector[n]     mu;   //means of responses
  mu = X * beta;
}
model{
  sigs ~ inv_gamma(0.01,0.01);         //variance
  beta ~ multi_normal(mu0,sigs*csq*V);
  for(i in 1:n){
    bsal[i] ~ normal(mu[i],sqrt(sigs));
  }
}
generated quantities {
  real Rbsq;
  Rbsq = 1 - sigs / v;
}"

#Format data for stan
stan_dat = list(n = n, v = v, V=solve(crossprod(X)), csq = n, X = X,
                mu0 = c(0,0,0,0,0), bsal = bank_salary$bsalary)

if (!file.exists("stanmodels/hw7q2.rda")){ #Generate the model if it is missing.
  mod_zellner = stan_model(model_code = stanmod_zellner)
  save(mod_zellner,file="stanmodels/hw7q2.rda",compress="xz")
}
load(file = "stanmodels/hw7q2.rda")
fit_zellner = sampling(mod_zellner,data=stan_dat,iter=100000,chains=4,
                       warmup = 50000,show_messages = FALSE);

#Format data for stan
stan_dat = list(n = n, V=solve(crossprod(X)), csq = 100^2, mu0 = c(0,0,0,0,0), 
                bsal = bank_salary$bsalary, X = X, v = v)

if (!file.exists("stanmodels/hw7q2.rda")){ #Generate the model if it is missing.
  mod_zellner = stan_model(model_code = stanmod_zellner)
  save(mod_zellner,file="stanmodels/hw7q2.rda",compress="xz")
}
load(file = "stanmodels/hw7q2.rda")
fit_100sq = sampling(mod_zellner,data=stan_dat,iter=100000,chains=4,
                       warmup = 50000,show_messages = FALSE);
```

All the questions below require you to use R and/or Stan to approximate the posterior distribution(s) of interest. I assume that you will do things like run multiple chains, calculate the Gelman-Rubin statistic, check the ACF of draws from the MCMC chain, determine the effective sample size, etc. However, I will only be looking at the posterior results. Make sure to include the R/Stan code used to run each model as an appendix so that I can check your work. I recommend a warmup of 50,000 iterations and 50,000 saved values for each parameter using 4 chains. \break
Ramsey and Schafer (2013) provide data related to bank employee salary for entrylevel clerical employees hired by the Harris Trust and Savings Bank between 1969 and 1977. Let the response variable be beginning salary (bsalary) in thousands of
dollars. Along with the intercept, there are four covariates:
* male, an indicator variable for whether the employee is male
* education, years of education
* experience, months of experience
* time, time at hiring as measured in months after January 1, 1969. 
The data are available in `bank_salary` data set in the **bayesutils** package available at https://github.com/jfrench/bayesutils. \break

Fit a multiple  linear regression model using all four explanatory variables and the typical normal error regression model. Use a Gamma(.01, .01) prior distribution for
the precision $\tau$ (or an Inv-Gamma(0.01, 0.01) prior distribution for the data variance $\sigma^2$). We will change the priors for $\beta$ in three different models below.

### Problem 1.
#### Model 1: Use independent normal priors for the regression coefficients, with mean 0 and variance 1002. Provide the posterior mean, standard deviation, 0.025, and 0.975 quantiles for all regression coefficients and the variance $\sigma^2$ for all three models in a simple tabular format. 

+----------+-----------+------------+------------+-----------+
|Parameter |Mean       |SD          |0.025q      |0.975q     |
+==========+===========+============+============+===========+
|$\beta_0$ |3.526321179|0.3313274714|2.874486e+00|4.176178439|
+----------+-----------+------------+------------+-----------+
|$\beta_1$ |0.722082593|0.1189208499|4.885812e-01|0.955499103|
+----------+-----------+------------+------------+-----------+
|$\beta_2$ |0.090059713|0.0249732311|4.083745e-02|0.139033216|
+----------+-----------+------------+------------+-----------+
|$\beta_3$ |0.001266128|0.0005962462|9.587489e-05|0.002436232|
+----------+-----------+------------+------------+-----------+
|$\beta_4$ |0.023419926|0.0052563613|1.302725e-02|0.033821225|
+----------+-----------+------------+------------+-----------+
|$\sigma^2$|0.263632159|0.0405537853|1.956672e-01|0.033697268|
+----------+-----------+------------+------------+-----------+
\newpage
### Problem 2.
#### Model 2: Use the default Zellner’s g-prior for $\beta$ ($c^2 = n$).  Provide the posterior mean, standard deviation, 0.025, and 0.975 quantiles for all regression coefficients and the variance $\sigma^2$ for all three models in a simple tabular format. 

+----------+-----------+------------+-------------+-----------+
|Parameter |Mean       |SD          |0.025q       |0.975q     |
+==========+===========+============+=============+===========+
|$\beta_0$ |3.488551155|0.4854034800|2.5346761457 |4.446111345|
+----------+-----------+------------+-------------+-----------+
|$\beta_1$ |0.715486239|0.1737947474|0.3745344344 |1.056050651|
+----------+-----------+------------+-------------+-----------+
|$\beta_2$ |0.089065187|0.0365580777|0.0169057889 |0.160682121|
+----------+-----------+------------+-------------+-----------+
|$\beta_3$ |0.001256217|0.0008687909|-0.0004515154|0.002963955|
+----------+-----------+------------+-------------+-----------+
|$\beta_4$ |0.023176023|0.0076808862|0.0081018553 |0.038304990|
+----------+-----------+------------+-------------+-----------+
|$\sigma^2$|0.571079268|0.0857885876|0.4273822269 |0.762295521|
+----------+-----------+------------+-------------+-----------+

### Problem 3.
#### Model 3: Use a conservative Zellner’s g-prior for $\beta$ with $c^2 = 1000$. Provide the posterior mean, standard deviation, 0.025, and 0.975 quantiles for all regression coefficients and the variance $\sigma^2$ for all three models in a simple tabular format. 

+----------+-----------+------------+------------+---------+
|Parameter |Mean       |SD          |0.025q      |0.975q   |
+==========+===========+============+============+=========+
|$\beta_0$ |3.525638710|0.3224330591|2.8923945423|4.1586533|
+----------+-----------+------------+------------+---------+
|$\beta_1$ |0.722118172|0.1164857662|0.4935365463|0.9516783|
+----------+-----------+------------+------------+---------+
|$\beta_2$ |0.090062717|0.0243253548|0.0422238087|0.1377916|
+----------+-----------+------------+------------+---------+
|$\beta_3$ |0.001267039|0.0005805062|0.0001321638|0.0024089|
+----------+-----------+------------+------------+---------+
|$\beta_4$ |0.023424966|0.0051279238|0.0133200260|0.0335112|
+----------+-----------+------------+------------+---------+
|$\sigma^2$|0.252165684|0.0376167956|0.1888764559|0.3360667|
+----------+-----------+------------+------------+---------+

\newpage
### Problem 4. 
#### Compare the posterior distributions for $R_b^2$ for all three models (with proper labeling) in a single plot. Which models appears to have the best fit? \break

The independent normal model and the Zellner's g-prior with $c^2 = 100^2$ both appear to be good fits. The g-prior with $c^2 = n$ is not a good fit.

```{r}
fits = rbind(cbind(subset(as.data.frame(fit),select='Rbsq'),
                   model = "Independent Normal"),
             cbind(subset(as.data.frame(fit_zellner),select='Rbsq'),
                   model = "G-Prior, c^2= n"),
             cbind(subset(as.data.frame(fit_100sq),select='Rbsq'),
                   model = "G-Prior, c^2 = 100^2"))
ggplot(fits, aes(x = Rbsq)) + theme_bw() +
  geom_density(aes(fill = model), alpha = 0.4)
```
\newpage

### Problem 5. 
#### Provide a point estimate (using the posterior means) of the regression equation for Model 1. Interpret each of the regression coefficients for Model 1 in the context of the problem. \break

Using the means from the independent normal model, we would expect a person's salary to follow this formula:

$$
S = 3.5 + 0.72I_m + 0.090E_Y + 0.0012X_M + 0.023T_M
$$
The average starting pay on January 1, 1969 for a woman with no experience or education was \$3,500. On average, men made \$720 more than women in the same position at the same time with equal experience and education. For each year of education, one could expect to earn \$90 more than their peers without an education, and one could expect to earn \$1.20 more for each month of experience. Wages steadily increased by an average of \$23 each month.

The dominating factor differentiating pay between peers in the sample was gender, with men making up to almost 20% more than their female peers and gender equating to eight years of education or 50 years of experience. Education is valued more than experience in wage calculation, with a year of education equating to six years and three months' worth of work experience. Average monthly wage increases are about twenty times those granted for work experience; assuming that wages were adjusted monthly for existing employees, one could expect to see an increase of \$24.20 per month.
\newpage

# Appendix

**Problem 1:**

```{r}
#imports
library(rstan); library(bayesutils); library(coda); 
library(ggplot2); library(bayesplot); 

#for use later
n = length(bank_salary$bsalary) #numentries
v = var(bank_salary$bsalary)    #sample variance

#stan
stanmod = "
data {
  int<lower=1>  n;    //num obs
  vector[n]     bsal; //main data
  vector[n]     male; //covariate
  vector[n]     educ; //covariate
  vector[n]     xp;   //covariate
  vector[n]     time; //covariate
  real<lower=0> v;    //sample variance
}
parameters{
  real<lower=0> sigs; //variance
  real          bet0; //regression coefficients vvv
  real          bet1;
  real          bet2;
  real          bet3;
  real          bet4;
}
model{
  //static priors* used by all models
  sigs ~ inv_gamma(0.01,0.01);//variance
  for(i in 1:n){              //data distribution
    bsal[i] ~ normal(bet0 + bet1*male[i] + bet2*educ[i] + bet3*xp[i] + bet4*time[i],sigs^(1./2));
  }
  
  //priors for regression coefficients; these change each time
  bet0 ~ normal(0,sqrt(1002)); 
  bet1 ~ normal(0,sqrt(1002));
  bet2 ~ normal(0,sqrt(1002));
  bet3 ~ normal(0,sqrt(1002));
  bet4 ~ normal(0,sqrt(1002));
}
generated quantities {
  real Rbsq;
  Rbsq = 1 - sigs / v;
}"

#Format data for stan
stan_dat = list(n = n, v = v, bsal = bank_salary$bsalary,
                male = bank_salary$male, educ = bank_salary$education, 
                xp = bank_salary$experience, time = bank_salary$time)

if (!file.exists("stanmodels/hw7q1.rda")){ #Generate the model if it is missing.
  mod = stan_model(model_code = stanmod)
  save(mod,file="stanmodels/hw7q1.rda",compress="xz")
}
load(file = "stanmodels/hw7q1.rda")
sink("NULL"); #keep from getting tons of output at once
fit = sampling(mod,data=stan_dat,iter=100000,chains=4,warmup = 50000,
               show_messages = FALSE);
sink(); #go back to printing

#Summarize
summary(fit)$summary
stan_dens(fit,par=c("bet0","bet1","bet2","bet3","bet4","sigs"),separate_chains=TRUE)
stan_trace(fit) + xlim(9.99e+04,1e+05)
stan_ac(fit)
posterior = as.array(fit)
mcmc_intervals(posterior,pars=c("bet0","bet1",'bet2',"bet3",'bet4',"sigs"))
samples = extract(fit)
ncycles = length(samples$sigs)
yrep = matrix(0,ncol=nrow(bank_salary),nrow=ncycles)
for(i in seq_len(nrow(bank_salary))){
  mui = {samples$bet0 + samples$bet1*bank_salary$male[i] + 
    samples$bet2*bank_salary$education[i] + 
    samples$bet3*bank_salary$experience[i] + samples$bet4*bank_salary$time[i]}
  yrep[,i] = rnorm(ncycles,mean=mui,sd=samples$sigs^(1./2))
}
ppc_hist(bank_salary$bsalary,yrep[sample(1:ncycles,8),])
```

\newpage

**Problem 2:**

```{r}
X = cbind(1, bank_salary$male, bank_salary$education, bank_salary$experience, 
          bank_salary$time)
stanmod_zellner = "
data {
  int<lower=1>  n;    //num obs
  vector[n]     bsal; //main data
  real<lower=0> csq;  //zellner's g-prior var
  matrix[n,5]   X;    //covariates
  vector[5]     mu0;  //starting means for covariants
  cov_matrix[5] V;    //covariance matrix
  real<lower=0> v;    //sample variance
}
parameters{
  real<lower=0> sigs; //variance
  vector[5]     beta; //regression coefficients
}
transformed parameters{
  vector[n]     mu;   //means of responses
  mu = X * beta;
}
model{
  sigs ~ inv_gamma(0.01,0.01);         //variance
  beta ~ multi_normal(mu0,sigs*csq*V);
  for(i in 1:n){
    bsal[i] ~ normal(mu[i],sqrt(sigs));
  }
}
generated quantities {
  real Rbsq;
  Rbsq = 1 - sigs / v;
}"

#Format data for stan
stan_dat = list(n = n, v = v, V=solve(crossprod(X)), csq = n, X = X,
                mu0 = c(0,0,0,0,0), bsal = bank_salary$bsalary)

if (!file.exists("stanmodels/hw7q2.rda")){ #Generate the model if it is missing.
  mod_zellner = stan_model(model_code = stanmod_zellner)
  save(mod_zellner,file="stanmodels/hw7q2.rda",compress="xz")
}
load(file = "stanmodels/hw7q2.rda")
sink("NULL"); #keep from getting tons of output at once
fit_zellner = sampling(mod_zellner,data=stan_dat,iter=100000,chains=4,
                       warmup = 50000,show_messages = FALSE);
sink(); #go back to printing

#Summarize
summary(fit_zellner,pars=c("beta","sigs"))$summary
stan_dens(fit_zellner,par=c("beta","sigs"),separate_chains=TRUE)
stan_trace(fit_zellner) + xlim(9.99e+04,1e+05)
stan_ac(fit_zellner)
posterior = as.array(fit_zellner)
#mcmc_intervals(posterior) #this one doesn't plot correctly
samples = extract(fit_zellner)
ncycles = length(samples$sigs)
yrep = matrix(0,ncol=nrow(bank_salary),nrow=ncycles)
for(i in seq_len(nrow(bank_salary))){
  mui = {mean(samples$beta[,1]) + mean(samples$beta[,2])*bank_salary$male[i] + 
    mean(samples$beta[,3])*bank_salary$education[i] + 
    mean(samples$beta[,4])*bank_salary$experience[i] + 
    mean(samples$beta[,5])*bank_salary$time[i]
  }
  yrep[,i] = rnorm(ncycles,mean=mui,sd=samples$sigs^(1./2))
}
ppc_hist(bank_salary$bsalary,yrep[sample(1:ncycles,8),])
```

\newpage

**Problem 3:**
```{r}
#Format data for stan
stan_dat = list(n = n, V=solve(crossprod(X)), csq = 100^2, mu0 = c(0,0,0,0,0), 
                bsal = bank_salary$bsalary, X = X, v = v)

if (!file.exists("stanmodels/hw7q2.rda")){ #Generate the model if it is missing.
  mod_zellner = stan_model(model_code = stanmod_zellner)
  save(mod_zellner,file="stanmodels/hw7q2.rda",compress="xz")
}
load(file = "stanmodels/hw7q2.rda")
sink("NULL"); #keep from getting tons of output at once
fit_100sq = sampling(mod_zellner,data=stan_dat,iter=100000,chains=4,
                       warmup = 50000,show_messages = FALSE);
sink(); #go back to printing

#Summarize
summary(fit_100sq,pars=c("beta","sigs"))$summary
stan_dens(fit_100sq,par=c("beta","sigs"),separate_chains=TRUE)
stan_trace(fit_100sq) + xlim(9.99e+04,1e+05)
stan_ac(fit_100sq)
posterior = as.array(fit_100sq)
#mcmc_intervals(posterior) #this one doesn't plot correctly
samples = extract(fit_100sq)
ncycles = length(samples$sigs)
yrep = matrix(0,ncol=nrow(bank_salary),nrow=ncycles)
for(i in seq_len(nrow(bank_salary))){
  mui = {mean(samples$beta[,1]) + mean(samples$beta[,2])*bank_salary$male[i] + 
    mean(samples$beta[,3])*bank_salary$education[i] + 
    mean(samples$beta[,4])*bank_salary$experience[i] + 
    mean(samples$beta[,5])*bank_salary$time[i]
  }
  yrep[,i] = rnorm(ncycles,mean=mui,sd=samples$sigs^(1./2))
}
ppc_hist(bank_salary$bsalary,yrep[sample(1:ncycles,8),])
```