---
title: "Homework 1 - Probability"
author: "Kitty Harris"
output:
  pdf_document: default
  html_notebook: default
---

### Problem 1: 
**Suppose that if $\theta = 1$, $y = N(5,2^2)$. If $\theta = 2$, then $y = \chi^2_3$. Assume $P(\theta = 1) = 0.3$ and $P(\theta = 2) = 0.7$.**

#### a. Determine the joint distribution of $y$ and $\theta$.

The joint distribution $p(y,\theta) = p(y|\theta)p(theta)$. We are already given $p(y|\theta = 1)$, $p(y|\theta = 2)$, $p(theta = 1)$, and $p(\theta) = 2$, and we know that $\theta$ can only take values 1 or 2. Therefore,

$$
p(y,\theta) = 
\begin{cases}
  \frac{3}{10}\frac{1}{\sqrt{8\pi}}\exp(-\frac{(y-5)^2}{4}) ,& \theta = 1 \\
  \frac{7}{10}\frac{2^{-3/2}}{\Gamma(3/2)}e^{-y/2}y^{1/2} ,& \theta = 2, y\geq0 \\
  0 ,& \theta=2,y<0
\end{cases}
$$

(The $\chi^2$ distribution has a non-negative support.)

#### b. Determine the marginal density of $y$.

The marginal density is the sum of the joint densities over each allowed value of $\theta$:

$$
p(y) = \frac{3}{10}\frac{1}{\sqrt{8\pi}}\exp(-\frac{(y-5)^2}{4}) + \frac{7}{10}\frac{2^{-3/2}}{\Gamma(3/2)}e^{-y/2}y^{1/2}I_{[0,\inf]}(y)
$$
\newpage <!-- keep plot with question -->

#### c. Draw the density of $y$.

Plot of $p(y)$ vs $y$:

```{r}
p_y = function(y){
  return(3/10*1/sqrt(8*pi)*exp(-(y-5)^2)/4 + 7/10*(2^(-3/2)/gamma(3/2))*exp(-y/2)*sqrt(y))
}
y = seq(0, 10, length.out=100)
plot(y,p_y(y))
```

#### d. Determine $P(\theta = 1 | y = 0.5)$.

Using $p(\theta|y) = \frac{p(\theta,y)}{p(y)}$:

$$p(\theta = 1 | y = 0.5) = \frac{p(\theta = 1, y = 0.5)}{p(y = 0.5)} = \frac{\frac{3}{10}\frac{1}{\sqrt{8\pi}}\exp(-\frac{(\frac{1}{2}-5)^2}{4})}{\frac{3}{10}\frac{1}{\sqrt{8\pi}}\exp(-\frac{(\frac{1}{2}-5)^2}{4}) + \frac{7}{10}\frac{2^{-3/2}}{\Gamma(3/2)}e^{-1/4}\frac{1}{2}^{1/2}}$$
```{r}
expterm = exp(-(1/2-5)^2/4) #for readability
coeff = 3/10*1/sqrt(8*pi)
total = coeff*expterm/(coeff*expterm + 7/10*2^-(3/2)/factorial(5/2)*exp(-1/4)*(1/2)^(1/2))
print(total)
```
\newpage <!-- keep answers with question -->

### Problem 2: 
**Suppose there are two species of panda bear. Both are equally common in the wild and live in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins 10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise birthing singleton infants. Assume these numbers are known with certainty, from many years of field research.**

#### a. What are the prior probabilities for each type of panda?

Since we do not know the relative probabilities of either species, we will begin by assuming that there is an even split in the population: $p(\theta = A) = p(\theta = B) = \frac{1}{2}$.

#### b. You observe a female panda of unkown species give birth to twins (with no other information). What is the probability that the female panda is species A?

The probability that it is of species A given that it gives birth to twins is $p(A|T) = \frac{P(y=T|\theta=A)p(\theta=A)}{p(y=T)}$. Finding $p(y=T)$ first,
$$
p(y=T) = \sum_\theta p(\theta)p(y=T|\theta) = \frac{1}{2}0.1+\frac{1}{2}0.2 = 0.15, \\
\therefore p(\theta = A | y = T) = \frac{0.1\frac{1}{2}}{0.15} = \frac{1}{3}.
$$

#### c. Assume the same female panda is about to give birth a second time. What is probability the second birth is also twins?

The probability that the second birth is also twins is 
$$
p(\tilde{y} = T | y = T) = \sum_\theta p(\tilde{y}=T|\theta)p(\theta|y=T) = 0.1\frac{1}{3} + 0.2\frac{2}{3} = \frac{1}{6}.
$$

#### d. Assume the female panda’s second birth is a singleton panda. In light of the two births (twins, then singleton), what is the probability that the female panda is species A?

The probability that the panda is of species A given this data is

$$
p(\theta = A | y = \{T,S\}) = \frac{p(y = \{T,S\}|\theta = A)p(\theta=A)}{p(y=T,S)}.
$$

The probability of the data given that the panda is of species A is 
$$
\begin{aligned}
p(y=\{T,S\} | \theta = A) &= p(y_1=T|\theta = A)p(y_2=S|\theta = A) \cr
                          &= p(y_1=T|\theta = A)(1-p(y_2=T|\theta = A) \cr
                          &= p(y_1=T|\theta = A)(1-p(y_1=T|\theta = A) = \frac{9}{100}. 
\end{aligned}
$$
Finding $p(y=\{T,S\})$ next,
$$
p(y=\{T,S\}) = p(y_1=T)(1-p(y_1=T)) = 0.1275,
$$

making $p(\theta = A | y = \{T,S\}) = \frac{0.09*0.5}{0.1275} = 0.35$.

\newpage <!-- keep answers with question -->

### Problem 3: Monty Hall
**Suppose you're on a game show, and you're given the choice of three doors: Behind one door is a car; behind the others, goats. You pick a door, say No. 1, and the host, who knows what's behind the doors, opens another door, say No. 3, which has a goat. He then says to you, "Do you want to pick door No.
2?" Is it to your advantage to switch your choice?**

#### a. What is the probability that you win assuming you switch doors?
The probability that we win assuming we switch doors is the same as $p(\theta = 2 | y = 3)$, the probability that the correct door was door No. 2 given that we were shown door No. 3. 
$$p(\theta = 2 | y = 3) = \frac{p(y = 3 | \theta = 2)p(\theta = 2)}{p(y = 3)}$$ 
Let our prior be $p(\theta = 1) = p(\theta = 2) = p(\theta = 3) = \frac{1}{3}$; before any doors are selected, each has an equal probability of being the correct door. The probability that we are shown door three, given door 2 is the winning door (and that we've selected door 1), is $p(y=3|\theta=2) = 1$; it's the only door that can be shown according to the rules of the game. The probability that we are shown door 3 is 
$$
p(y=3) = \sum_\theta p(y=3|\theta)p(\theta) = \frac{1}{2}\frac{1}{3} + 1\frac{1}{3} + 0\frac{1}{3} = \frac{1}{2}.
$$
Therefore, the probability of winning if we switch is
$$
p(\theta = 2 | y = 3) = \frac{1\frac{1}{3}}{\frac{1}{2}} = \frac{2}{3}.
$$

#### b. What is the probability that you win assuming you don’t switch doors?
The probability that we win assuming we do not switch doors is 
$$
p(\theta = 1 | y = 3) = \frac{p(y=3|\theta=1)p(\theta=2)}{p(y=3)}.
$$
Since either remaining door can be opened if we've already selected the winning one, $p(y=3|\theta=1)=\frac{1}{2}$. Therefore, the probability that we win without switching doors is 
$$
p(\theta = 1 | y = 3) = \frac{\frac{1}{2}\frac{1}{3}}{\frac{1}{2}} = \frac{1}{3}.
$$