---
title: "Homework5"
author: "Kitty Harris"
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

### Problem 1. 
**Assume $y_1,y_2,...y_100 \iid InvGamma(\alpha,\beta)$ with $\alpha=3$. Assume $\beta ~ Gamma(4,4)$. Use the Metropolis-Hastings algorithm to approximate the posterior distribution of $\beta$.**

#### a. Generate some observed data $y$. To do this, execute the following commands:
```{r}
library(bayesutils)
set.seed(77) 
y = bayesutils::rinvgamma(100, 2, 3)
```
**Double-check that you get the correct data by executing the following commands to make sure you get the same answers.**
```{r}
range(y) #[1] 0.4647718 12.0184742 
mean(y) #[1] 2.520916 
sd(y) #[1] 2.244544 
range(y) #[1] 0.4647718 12.0184742
```

#### b. Construct a plot of the unnormalized posterior density of $\beta$. Tip: First compute this on the log scale for a sequence of $\beta$ values. Then subtract the largest value (still on the log scale) and transform the values back to the original scale with the exponential function.


#### c. Decide on a proposal distribution for $\beta$. Construct a plot of the proposal distribution. Compare this to the previous plot. Repeat the process until they have at least somewhat similar shapes. What proposal distribution did you decide on?

#### d. Run 5 MCMC chains with a range of starting values for at least 100,000 iterations. Discard the first half of each chain as warmup. Then combine the results into a mcmc.list and use the summary function to summarize the results.

#### e. Assess convergence of your chains.

### Problem 2.
**Assume $y_1,y_2,...y_100 \iid InvGamma(\alpha,\beta)$ with $\beta=4$. Assume $\alpha ~ ????(1,5)$. Use the Metropolis-Hastings algorithm to approximate the posterior distribution of $\alpha$. Assume the data are the same as from the previous problem.**

#### a. Construct a plot of the unnormalized posterior density of $\alpha$.

#### b. Decide on a proposal distribution for $\alpha$. Construct a plot of the proposal distribution. Compare this to the previous plot. Repeat the process until they have at least somewhat similar shapes. What proposal distribution did you decide on?

#### c. Run 5 MCMC chains with a range of starting values for at least 100,000 iterations. Discard the first half of each chain as warmup. Then combine the results into a mcmc.list and use the summary function to summarize the results.

#### d. Assess convergence of your chains.