---
title: "Hw08- Categorical Predictors"
subtitle: Math 7393 Bayesian Statistics
output:
  pdf_document: default
---

All of the questions below require you to use R and/or Stan to approximate the posterior distribution(s) of interest.  I assume that you will do things like run multiple chains, calculate the Gelman-Rubin statistic, check the ACF of draws from the MCMC chain, determine the effective sample size, etc.  However, I will only be looking at the posterior results.  I recommend a warmup of 50,000 iterations and 50,000 saved values for each parameter using 4 chains. Please include your Stan model code as an appendix that I can look at if necessary, but not in the actual questions. One way to do this in R Markdown/Quarto is to set the `include` or `echo` option of the code chunk to `FALSE`/`false` so that it runs but is not included in the output. Then manually paste the relevant code to an appendix between two sets of 3 back ticks like you are starting and concluding a code chunk, but don't include the `{r}` part in the starting part.

**Hawks data**

Load the `Hawks` data set from the **Stat2Data** package. The data are a sample of 908 hawks from three different species. More details can be found by running `?Stat2Data::Hawks`. The variables in the data set include:

- `Species`:	`CH`=Cooper's, `RT`=Red-tailed, `SS`=Sharp-Shinned.
- `Wing`: length (mm) of primary wing feather from tip to wrist it attaches to.
- `Weight`:	body weight (g).

Run the following commands to load this data set and remove any `NA` values.

```{r}
data(Hawks, package = "Stat2Data")
Hawks <- na.omit(Hawks[,c("Species", "Wing", "Weight")])
```

We will fit four different regression models to the data using `Wing` as the response variable and the other variables as explanatory variables.

*Data distribution*

$y_i \mid \mu_i, \sigma^2 \sim N(\mu_i, \sigma^2)$ with $y_1, y_2, \ldots, y_n$ being independent and $\mu_i$ is based on the appropriate regression model (like in class).

*Prior distributions*

$\sigma^2 \sim$ Inv-Gamma(0.01, 0.01).
All regression coefficients are assumed to be independent and have a $N(0,100^2)$ prior distribution.

Model 1: A one-way ANOVA using `species` as the categorical predictor and "CH" as the reference level.
Model 2: A simple linear regression model using `Weight` as the predictor.
Model 3: A parallel lines (main effects) model using `Weight` as the quantitative predictor and `species` as the categorical predictor with "CH" as the reference level.
Model 4: A separate lines (interaction) model using `Weight` as the quantitative predictor and `species` as the categorical predictor with "CH" as the reference level.

# Problem 1 

Create a scatter plot of the data using `Wing` as the y variable, `Weight` as the x variable, and distinguishing between the `species` using appropriate symbols/colors.  Include the appropriate legend. Do you notice a relationship between the three variables?

```{r}
CH = Hawks[which(Hawks$Species == "CH"), ]
RT = Hawks[which(Hawks$Species == "RT"), ]
SS = Hawks[which(Hawks$Species == "SS"), ]
plot(Hawks$Weight,Hawks$Wing,xlab="Weight (g)",ylab="Wingspan (mm)") #Get axes right
points(CH$Weight,CH$Wing,col='blue')
points(RT$Weight,RT$Wing,col='red')
points(SS$Weight,SS$Wing,col='green')

legend(1500,200,c("Cooper's","Red-Tailed","Sharp-Shinned"),col=c('blue','red','green'),pch='o')
```

There is a positive relationship between weight and wingspan. Species limits the range of wingspans to an extent, although there are definite outliers. The Sharp-Shinned hawks are the smallest of the three species while the Red-Tailed are the largest. There also appear to be two possibly distinct size groups within the Cooper's and Sharp-Shinned hawk populations.

# Problem 2

Fit a one-way ANOVA model (Model 1) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

```{r include=FALSE}
library(rstan);
anova = "
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  //vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  //real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + alpha1*IR + alpha2*IS;
}
model{
    //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  //beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
"

if (!file.exists("stanmodels/hw8anova.rda")){
  anova_mod = stan_model(model_code = anova)
  save(anova_mod,file = "stanmodels/hw8anova.rda",compress="xz")
}
load("stanmodels/hw8anova.rda")
dat = list(n = length(Hawks$Wing),v = var(Hawks$Wing),
           Wing = Hawks$Wing, #Weight = Hawks$Weight,
           IR= as.integer(Hawks$Species == "RT"),
           IS= as.integer(Hawks$Species == "SS"))
fit_anova = sampling(anova_mod,data=dat,iter=50000,chains=4,cores=4)
summary(fit_anova,c("beta0","alpha1","alpha2","Rbsq","sigs"),use_cache=FALSE)$summary
```

+----------+-----+-----+-----+
|Parameter |0.025|mean |0.975|
+==========+=====+=====+=====+
|$\beta_0$ |237  |244  |251  |
+----------+-----+-----+-----+
|$\alpha_1$|132  |140. |147  |
+----------+-----+-----+-----+
|$\alpha_2$|-66.7|-59.1|-51.4|
+----------+-----+-----+-----+
|$\sigma^2$|770. |845  |927  |
+----------+-----+-----+-----+

# Problem 3

Provide a point estimate (using the posterior means) of the regression equation for the one-way ANOVA model (Model 1). Interpret each coefficient in the context of the problem.

Using the means, a point estimate for the regression equation of the wingspan, in mm, would be:

$$ \mathrm{Wingspan} = 244 + 140I_{\mathrm{Red-Tailed}} - 59.1I_{\mathrm{Sharp-Shinned}} $$
Interpretation: The average wingspan for a Cooper's hawk is 244 mm, while the average wingspan for a Red-Tailed hawk is 140 mm more and that of a Sharp-Shinned hawk is 59.1 mm less.

# Problem 4

Fit a simple linear regression model (Model 2) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

```{r include=FALSE}
library(rstan)
linear = "
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  //vector[n]    IR;      //indicator for Red-Tailed
  //vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  //real          alpha1; //response for red-tailed
  //real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + beta1*Weight;
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  //alpha1 ~ normal(0,100);
  //alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
"

if (!file.exists("stanmodels/hw8linear.rda")){
  linear_mod = stan_model(model_code = linear)
  save(linear_mod,file = "stanmodels/hw8linear.rda",compress="xz")
}
load("stanmodels/hw8anova.rda")
dat = list(n = length(Hawks$Wing),v = var(Hawks$Wing),
           Wing = Hawks$Wing, Weight = Hawks$Weight#,
           #IR= as.integer(Hawks$Species == "RT"),
           #IS= as.integer(Hawks$Species == "SS")
           )
fit_linear = sampling(linear_mod,data=dat,iter=50000,chains=4,cores=4)
summary(fit_linear,c("beta0","beta1","Rbsq","sigs"),use_cache=FALSE)$summary
```

+----------+-----+-----+-----+
|Parameter |0.025|mean |0.975|
+==========+=====+=====+=====+
|$\beta_0$ |163  |167  |171  |
+----------+-----+-----+-----+
|$\beta_1$ |0.188|0.193|0.197|
+----------+-----+-----+-----+
|$\sigma^2$|1046 |1147 |1258 |
+----------+-----+-----+-----+

# Problem 5

Provide a point estimate (using the posterior means) of the regression equation for the simple linear regression model (Model 2). Interpret each coefficient in the context of the problem.

Using the means, a point estimate for the regression equation of the wingspan, in mm, would be:

$$ \mathrm{Wingspan} = 167 + 0.193\mathrm{Weight}$$,

where weight is given in g.

Interpretation: Each gram of weight corresponds to, on average, a 0.193 mm increase in wingspan. There is an additional offset of 167 mm. For example, the wingspan of a "weightless" hawk (if hawks could be weightless) would be 167 mm, and the wingspan of a hawk weighing 1000 g would be 360 mm.


# Problem 6

Fit a parallel lines regression model (Model 3) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

```{r include=FALSE}
library(rstan); library(dplyr)
parallel = "
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + beta1*Weight + alpha1*IR + alpha2*IS;
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  //log likelihood
  real log_lik[n];
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
"

if (!file.exists("stanmodels/hw8parallel.rda")){
  parallel_mod = stan_model(model_code = parallel)
  save(parallel_mod,file = "stanmodels/hw8parallel.rda",compress="xz")
}
load("stanmodels/hw8parallel.rda")
dat = list(n = length(Hawks$Wing),v = var(Hawks$Wing),
           Wing = Hawks$Wing, Weight = Hawks$Weight,
           IR= as.integer(Hawks$Species == "RT"),
           IS= as.integer(Hawks$Species == "SS"))
fit_parallel = sampling(parallel_mod,data=dat,iter=50000,chains=4,cores=4)
summary(fit_parallel,c("beta0","beta1","alpha1","alpha2","Rbsq","sigs"),use_cache=FALSE)$summary
```

+----------+------+------+------+
|Parameter |0.025 |mean  |0.975 |
+==========+======+======+======+
|$\beta_0$ |206   |213   |221   |
+----------+------+------+------+
|$\beta_1$ |0.0620|0.0727|0.0833|
+----------+------+------+------+
|$\alpha_1$|80.9  |90.6  |100.  |
+----------+------+------+------+
|$\alpha_2$|-47.0 |-39.4 |-31.8 |
+----------+------+------+------+
|$\sigma^2$|644   |706   |774   |
+----------+------+------+------+

# Problem 7

Provide a point estimate (using the posterior means) of the regression equation for the parallel lines regression model (Model 3). Interpret each coefficient related to the `CH` and `RT` species in the context of the problem.

Using the means, a point estimate for the regression equation of the wingspan, in mm, would be:

$$ \mathrm{Wingspan} = 213 + 0.0727\mathrm{Weight} + 90.6I_{\mathrm{Red-Tailed}} - 39.4I_{\mathrm{Sharp-Shinned}}$$
Interpretation: The base mean wingspan for a Cooper's hawk is 213 mm, while on average a Red-Tailed hawk's wingspan will be 90.6 mm greater and a Sharp-Shinned hawk's wingspan will be 39.4 less. For each additional gram the hawk weighs, we expect to see 0.0727 mm more of wingspan, regardless of the hawk's species.

# Problem 8

Fit a separate lines regression model (Model 4) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

```{r include=FALSE}
library(rstan)
separate = "
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
  real          delta1; //response by weight for red-tailed
  real          delta2; //response by weight for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  for(i in 1:n){
    mu[i] = beta0 + beta1*Weight[i] + (alpha1 + delta1*Weight[i])*IR[i] + (alpha2 + delta2*Weight[i])*IS[i];
  }
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  delta1 ~ normal(0,100);
  delta2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){
    Wing[i] ~ normal(mu[i],sqrt(sigs));
  }
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
"

if (!file.exists("stanmodels/hw8sep.rda")){
  separate_mod = stan_model(model_code = separate)
  save(separate_mod,file = "stanmodels/hw8sep.rda",compress="xz")
}
load("stanmodels/hw8sep.rda")
dat = list(n = length(Hawks$Wing),v = var(Hawks$Wing),
           Wing = Hawks$Wing, Weight = Hawks$Weight,
           IR= as.integer(Hawks$Species == "RT"),
           IS= as.integer(Hawks$Species == "SS"))
fit_separate = sampling(separate_mod,data=dat,iter=50000,chains=4,cores=4)
summary(fit_separate,c("beta0","beta1","alpha1","alpha2","delta1","delta2","Rbsq","sigs"),use_cache=FALSE)$summary
```
+----------+-------+-------+-------+
|Parameter |0.025  |mean   |0.975  |
+==========+=======+=======+=======+
|$\beta_0$ |181    |198    |214    |
+----------+-------+-------+-------+
|$\beta_1$ |0.0736 |0.110  |0.147  |
+----------+-------+-------+-------+
|$\alpha_1$|104    |125    |145    |
+----------+-------+-------+-------+
|$\alpha_2$|-66.2  |-48.6  |-30.9  |
+----------+-------+-------+-------+
|$\delta_1$|-0.0926|-0.0541|-0.0161|
+----------+-------+-------+-------+
|$\delta_2$|0.0777 |0.131  |0.183  |
+----------+-------+-------+-------+
|$\sigma^2$|587    |665    |707    |
+----------+-------+-------+-------+

# Problem 9

Provide a point estimate (using the posterior means) of the regression equation for the parallel lines regression model (Model 4). Interpret each coefficient related to the `CH` and `RT` species in the context of the problem.

Using the means, a point estimate for the regression equation of the wingspan, in mm, would be:

$$ \mathrm{Wingspan} = 198 + 0.11\mathrm{Weight} + (125 - 0.0541\mathrm{Weight})I_{\mathrm{Red-Tailed}} + (-48.6 + 0.131\mathrm{Weight})I_{\mathrm{Sharp-Shinned}} $$

Interpretation for Cooper's and Red-Tailed: The base wingspan, before accounting for weight, of a Cooper's hawk is 198 mm, and the Red-Tailed hawk's is 125 mm more (323 mm) before accounting for weight. For each gram of weight, the Cooper's hawk's wingspan increases by 0.11 mm, while the Red-Tailed hawk's increases by 0.0541 less (for a net of 0.559 mm/g), or roughly have as much. Therefore, while the Red-Tailed hawks are generally larger than the Cooper's hawks, the Cooper's hawks' weights more strongly drive their wingspans than the Red-Tailed hawks'.

# Problem 10

Compute the WAIC and LOOIC of each the 4 models, summarizing your results in a simple table.  Which model do you think is most appropriate?  Why? 

```{r include=FALSE}
library(loo)
#anova
ll_anova = extract_log_lik(fit_anova,merge_chains=FALSE)
waic_anova = waic(ll_anova)
looic_anova = loo(ll_anova,r_eff = exp(relative_eff(ll_anova)))
#linear
ll_linear = extract_log_lik(fit_linear,merge_chains=FALSE)
waic_linear = waic(ll_linear)
looic_linear = loo(ll_linear,r_eff = exp(relative_eff(ll_linear)))
#parallel
ll_parallel = extract_log_lik(fit_parallel,merge_chains=FALSE)
waic_parallel = waic(ll_parallel)
looic_parallel = loo(ll_parallel,r_eff = exp(relative_eff(ll_parallel)))
#separate
ll_separate = extract_log_lik(fit_separate,merge_chains=FALSE)
waic_separate = waic(ll_separate)
looic_separate = loo(ll_separate,r_eff = exp(relative_eff(ll_separate)))

loo_compare(waic_anova,waic_linear,waic_parallel,waic_separate)
loo_compare(looic_anova,looic_linear,looic_parallel,looic_separate)
```
+--------+----+-----+
|Model   |WAIC|LOOIC|
+========+====+=====+
|ANOVA   |8613|8613 |
+--------+----+-----+
|Linear  |8883|8883 |
+--------+----+-----+
|Parallel|8468|8466 |
+--------+----+-----+
|Separate|8399|8397 |
+--------+----+-----+

The model with the best WAIC and LOOIC results is the separate lines model. It also appears to be the best fit visually, as the three species all clearly have different wingspan ranges but also do not appear to have the same relationships as each other between weight and wingspan. However, while the separate line model is the best of these options, the WAIC and LOOIC criteria are extremely poor for all of the models. Because other indicators of convergence pass, I believe this is because of one of the following:

1. Log-Likelihood, which is not used by other indicators, is defined incorrectly
2. Sample variance is too high.

However, I was not able to successfully diagnose the issue before the submission date.

\newpage

# Appendix

**ANOVA**
```
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  //vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  //real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + alpha1*IR + alpha2*IS;
}
model{
    //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  //beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
```

**Normal Linear Regression**
```
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  //vector[n]    IR;      //indicator for Red-Tailed
  //vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  //real          alpha1; //response for red-tailed
  //real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + beta1*Weight;
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  //alpha1 ~ normal(0,100);
  //alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
```

**Parallel Lines**
```
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + beta1*Weight + alpha1*IR + alpha2*IS;
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){Wing[i] ~ normal(mu[i],sqrt(sigs));}
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  //log likelihood
  real log_lik[n];
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
```

**Separate Lines**
```
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
  real          delta1; //response by weight for red-tailed
  real          delta2; //response by weight for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  for(i in 1:n){
    mu[i] = beta0 + beta1*Weight[i] + (alpha1 + delta1*Weight[i])*IR[i] + (alpha2 + delta2*Weight[i])*IS[i];
  }
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
  delta1 ~ normal(0,100);
  delta2 ~ normal(0,100);
  
  //data dist
  for(i in 1:n){
    Wing[i] ~ normal(mu[i],sqrt(sigs));
  }
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_lik[n]; //log likelihood
  for(i in 1:n){log_lik[i] = normal_lpdf(Wing[i] | mu[i],sqrt(sigs));}
}
```