---
title: "Hw08- Categorical Predictors"
subtitle: Math 7393 Bayesian Statistics
output:
  pdf_document: default
---

All of the questions below require you to use R and/or Stan to approximate the posterior distribution(s) of interest.  I assume that you will do things like run multiple chains, calculate the Gelman-Rubin statistic, check the ACF of draws from the MCMC chain, determine the effective sample size, etc.  However, I will only be looking at the posterior results.  I recommend a warmup of 50,000 iterations and 50,000 saved values for each parameter using 4 chains. Please include your Stan model code as an appendix that I can look at if necessary, but not in the actual questions. One way to do this in R Markdown/Quarto is to set the `include` or `echo` option of the code chunk to `FALSE`/`false` so that it runs but is not included in the output. Then manually paste the relevant code to an appendix between two sets of 3 back ticks like you are starting and concluding a code chunk, but don't include the `{r}` part in the starting part.

**Hawks data**

Load the `Hawks` data set from the **Stat2Data** package. The data are a sample of 908 hawks from three different species. More details can be found by running `?Stat2Data::Hawks`. The variables in the data set include:

- `Species`:	`CH`=Cooper's, `RT`=Red-tailed, `SS`=Sharp-Shinned.
- `Wing`: length (mm) of primary wing feather from tip to wrist it attaches to.
- `Weight`:	body weight (g).

Run the following commands to load this data set and remove any `NA` values.

```{r}
data(Hawks, package = "Stat2Data")
Hawks <- na.omit(Hawks[,c("Species", "Wing", "Weight")])
```

We will fit four different regression models to the data using `Wing` as the response variable and the other variables as explanatory variables.

*Data distribution*

$y_i \mid \mu_i, \sigma^2 \sim N(\mu_i, \sigma^2)$ with $y_1, y_2, \ldots, y_n$ being independent and $\mu_i$ is based on the appropriate regression model (like in class).

*Prior distributions*

$\sigma^2 \sim$ Inv-Gamma(0.01, 0.01).
All regression coefficients are assumed to be independent and have a $N(0,100^2)$ prior distribution.

Model 1: A one-way ANOVA using `species` as the categorical predictor and "CH" as the reference level.
Model 2: A simple linear regression model using `Weight` as the predictor.
Model 3: A parallel lines (main effects) model using `Weight` as the quantitative predictor and `species` as the categorical predictor with "CH" as the reference level.
Model 4: A separate lines (interaction) model using `Weight` as the quantitative predictor and `species` as the categorical predictor with "CH" as the reference level.

# Problem 1 

Create a scatter plot of the data using `Wing` as the y variable, `Weight` as the x variable, and distinguishing between the `species` using appropriate symbols/colors.  Include the appropriate legend. Do you notice a relationship between the three variables?

```{r}
CH = Hawks[which(Hawks$Species == "CH"), ]
RT = Hawks[which(Hawks$Species == "RT"), ]
SS = Hawks[which(Hawks$Species == "SS"), ]
plot(Hawks$Weight,Hawks$Wing,xlab="Weight (gm)",ylab="Wingspan (mm)") #Get axes right
points(CH$Weight,CH$Wing,col='blue')
points(RT$Weight,RT$Wing,col='red')
points(SS$Weight,SS$Wing,col='green')

legend(1500,200,c("Cooper's","Red-Tailed","Sharp-Shinned"),col=c('blue','red','green'),pch='o')
```

There is a positive relationship between weight and wingspan. Species limits the range of wingspans to an extent, although there are definite outliers. The Sharp-Shinned hawks are the smallest of the three species while the Red-Tailed are the largest. There also appear to be two possibly distinct size groups within the Cooper's and Sharp-Shinned hawk populations.

# Problem 2

Fit a one-way ANOVA model (Model 1) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

```{r include=FALSE}
library(rstan); library(dplyr)
anova = "
data{
  int<lower=1> n;       //num observations
  vector[n]    Wing;    //wingspans (RESPONSE VARIABLE)
  vector[n]    Weight;  //weights
  vector[n]    IR;      //indicator for Red-Tailed
  vector[n]    IS;      //indicator for Sharp-Shinned
  real<lower=0> v;      //sample variance of wingspans
}
parameters{
  real<lower=0> sigs;   //sigma squared
  real          beta0;  //baseline
  real          beta1;  //response by weight
  real          alpha1; //response for red-tailed
  real          alpha2; //response for sharp-shinned
}
transformed parameters{
  vector[n]     mu;     //mean of responses
  mu = beta0 + beta1*Weight + alpha1*IR + alpha2*IS;
}
model{
  //priors:
  sigs   ~ inv_gamma(0.01,0.01);
  beta0  ~ normal(0,100);
  beta1  ~ normal(0,100);
  alpha1 ~ normal(0,100);
  alpha2 ~ normal(0,100);
}
generated quantities{
  real Rbsq;
  Rbsq = 1 - sigs/v;
  
  real log_ll[n]; //log likelihood PER observation:
  for (i in 1:n) log_ll[i] = normal_lpdf(Wing[i] | mu[i], sqrt(sigs));
}
"

if (!file.exists("stanmodels/hw8anova.rda")){
  anova_mod = stan_model(model_code = anova)
  save(anova_mod,file = "stanmodels/hw8anova.rda",compress="xz")
}
load("stanmodels/hw8anova.rda")
dat = list(n = length(Hawks$Wing),v = var(Hawks$Wing),
           Wing = Hawks$Wing, Weight = Hawks$Weight,
           IR= as.integer(Hawks$Species == "RT"),
           IS= as.integer(Hawks$Species == "SS"))
fit_anova = sampling(anova_mod,data=dat,iter=100000,chains=4,cores=4)
summary(fit_anova,"Rbsq",use_cache=FALSE)
```
# Problem 3

Provide a point estimate (using the posterior means) of the regression equation for the one-way ANOVA model (Model 1). Interpret each coefficient in the context of the problem.

# Problem 4

Fit a simple linear regression model (Model 2) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

# Problem 5

Provide a point estimate (using the posterior means) of the regression equation for the simple linear regression model (Model 2). Interpret each coefficient in the context of the problem.


# Problem 6

Fit a parallel lines regression model (Model 3) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

# Problem 7

Provide a point estimate (using the posterior means) of the regression equation for the parallel lines regression model (Model 3). Interpret each coefficient related to the `CH` and `RT` species in the context of the problem.

# Problem 8

Fit a separate lines regression model (Model 4) using Stan. In a simple, tabular format, provide the posterior mean, the 0.025 quantile, and the 0.975 quantile for all regression coefficients and the variance $\sigma^2$.

# Problem 9

Provide a point estimate (using the posterior means) of the regression equation for the parallel lines regression model (Model 4). Interpret each coefficient related to the `CH` and `RT` species in the context of the problem.

# Problem 10

Compute the WAIC and LOOIC of each the 4 models, summarizing your results in a simple table.  Which model do you think is most appropriate?  Why? 

\newpage

# Appendix

##Problem 2
```

```